Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-29 article/main.tex
fix typo
les chiffres signicatifs
refix typo
3 tonnes de TODO pour floette

    \draw (0.3, 1.1) -- (0.3, 1.5) -- (-1, 1.5) -- (-1, -4) -- (5.8, -4);
    \draw (-1.1, -2.9) -- +(0.2, -0.2);
    \path (-0.9, -3) node[anchor=west] {6};

    \draw[->] (x) -- +(0, -1.5);
    \node at (6.4, -6.25) (xi) {$X_i$};
    \draw[->] (6.4, -4.4) -- (xi);

@@

@@

Let $\LCG_{n,k}(x, c)$ denote the vector $(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they are obvious from the context).  It is easy to check that:
\begin{align}
\label{eq:lcg-additive}
\LCG_{n,k}(x + y, c + d) &= \LCG_{n,k}(x, c) + \LCG_{n,k}(y, d),  \\
\label{eq:lcg-scalar}
\LCG_{n,k}(\lambda x, \lambda c) &= \lambda \LCG_{n,k}(x, c).   
\end{align}
Hence those sets of vector have a lot of structure we can use.

The LCG have been largely studied. This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}. A particular LCG is the Lehmer's generator \cite{Lehmer}, where the multiplier \(a\) is known and the increment is zero.

@@

\begin{itemize}
	\item By \ref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
	\item By \ref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
\end{itemize}

@@

\[\begin{pmatrix}
1&a&a^2&\dots&a^{n-1}\\
0&2^k&0&\dots&0\\
0&0&2^k&\dots&0\\
\dots&\dots&\dots&\dots&\dots\\
0&0&0&\dots&2^k\\
\end{pmatrix}\]

@@

We want to find an element \(U\) in the lattice only knowing the most significant bits of each of its coordinates.

We decompose \(U = T + (T-U)\) where \(T\) contains only the upper bits and \(T-U\) the lowers one. There is no reason for \(T\) to be in the lattice, but as \(T-U\) is small, we know \(T\) is \emph{close} to \(U\).

@@

In this article, we will only use an exact CVP solver when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator. Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{122-\ell} \). So, if we fix \(\ell\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122-\ell} \leqslant \lambda_1(L(128-\ell,n))\].

@@

As using an exact CVP solver is costly, we'll try a cheaper but less precise way to find the closest vector.

We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).

\medskip

@@

In this article, we will only use the Babai's rounding when \(k = 64\) and when we know the \(6+\ell\) upper bit of our Lehmer's generator. Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{58-\ell} \). So, if we fix \(n\) we can search for the minimum \(\ell\) such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{58-\ell} \leqslant \lambda_1(L(n,64))\]

\begin{table}[htbp!]
  \centering
  \begin{tabular}{|c|c|c|c|}
	\hline
	n & \(\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert\) & \( \lambda_1(L(n,64)) \) & minimum \(\ell\) \\
	\hline
	3 & 2.869065991049632 & \(4.08909120094543e^{12} \simeq 2^{42}\) & 19 \\
	4 & 2.060717188416003 & \(2.43569932844650e^{14} \simeq 2^{48}\) & 13 \\
	5 & 3.7692142154392747 & \(1.71669732658884e^{15}\simeq 2^{51}\) & 11 \\
	6 & 2.6925925483728537 & \(1.02550589786192e^{16}\simeq 2^{53}\) & 8 \\
	\hline
  \end{tabular}
  \caption{\todo{WRITE-ME}}
  \label{tab:babai}
\end{table}

@@

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations give access to $S_i[58:64+\ell]$ for the corresponding
states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
linear congruential generator on $64$ bits, where we have access to the most
$6+\ell$ bits of each state (denoted by $T$), for a few consecutive states. This
is sufficient to reconstruct entirely the successive states of this truncated
linear congruential generator. This reveals $S_0[\ell:64+\ell]$, and
using~\eqref{eq:output} the entire $S_0$ can be reconstructed. The precise
details follow.

@@

We consider the first three internal states
$\mathbf{S} = (S_0, S_1, S_2) = \LCG_{3, 128}(S_0, c)$. We will guess the $\ell$
least-significant bits of $S_0$, therefore let us assume that their value is
known and denote it by $w$. We define $\mathbf{S}' = \LCG_{3, 128}(S_0 - w,
0)$. By~\eqref{eq:lcg-additive}, we have
$\mathbf{S}' = \mathbf{S} - \LCG_{3, 128}(w, c)$. The point is that the elements
of $\mathbf{S}'$ follow a geometric progression of common ratio $a$; in
addition, the $\ell$ least significant bits of each components are zero. The
crux of the reconstruction algorithm is to find $\mathbf{S}'$.

For this, set $\mathbf{T}' = \left(\mathbf{S}' / 2^\ell\right) \bmod 2^{64}$
(the division by $2^\ell$ is exact). It follows that the $\mathbf{T}'$ is also a
geometric progression with common ratio $a$, and we have access to the $\ell+6$
high order bits of $T'_i$ if we guess the $i$-th rotation. We have seen in
section~\ref{sec:geometric} that 19 known bits on 3 consecutive outputs are
sufficient for a deterministic reconstruction of the full sequence. Therefore,
we guess $\ell = 19$ low-order bits of the state and three consecutive
rotations.

Therefore, the algorithm that reconstructs the internal state of the
\textsf{PCG64} generator with known increment proceeds as shown in algorithm~\ref{algo:known}.

\begin{algorithm}
\begin{algorithmic}[1]
    \Procedure{ReconstructState${}_\ell$}{$X_0, X_1, X_2$}
    \For{$0 \leq w < 2^{\ell}$} \Comment{Guess least-significant bits of $S_0$}
    \State $w_j = a^j w \bmod 2^{\ell}$ \Comment{Propagate to $S_1, S_2$}
    \State $S'_j \gets a^j w + c \sum_{k = 0}^j a^k \bmod 2^{128}$    \Comment{Shifted states}

@@

    \State $ T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$  \Comment{Truncated LCG output}

    \State $ T'_j \gets    T_j -  S'_i[58:64+\ell]$  \Comment{Shifted Truncated LCG output}

    \State $(U_0, U_1, U_2) \gets \left\lfloor (T'_0, T'_1, T'_2) \cdot 2^{58-\ell} \cdot \widetilde G_3^{-1} \right\rceil \cdot \widetilde G_3$ \Comment{CVP via Babai rounding}

    \State $ S_0[0:64] \gets  S'_0[0:64] + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
    \State $ S_0[64:128] \gets S_0[0:64] \oplus Y_0$

    \State $ S_1 \gets a  S_0 + c$ \Comment{Recompute $X_1$}
    \State $\widehat{Y}_1 =  S_1[0:64] \oplus  S_1[64:128]$

    \If{$\widehat {Y}_1 = Y_1$} \Comment{Check consistency}
    \State output $ S_0$ as a candidate internal state.
    \EndIf
    \EndFor
    \EndFor
    \EndProcedure
  \end{algorithmic}
  \caption{State reconstruction Algorithm (case where $c$ is known)}
  \label{algo:known}
\end{algorithm}

% \begin{theorem}
%   Algorithm \textsc{ReconstructState} is correct (it always return the correct internal state).
% \end{theorem}

% \todo{avec la nouvelle partie 3 et le blabla au dessus de l'algo, y'a plus vraiment besoin de preuve}
% Charles: je suis plutôt d'accord.

@@

The algorithm of section~\ref{sec:Cknown} does not apply directly to the general
case where the value of $c$ is unknown. A ``baseline'' procedure would consist
in guessing $S_0[64:128]$ and $S_1[64:128]$; using eq.~\eqref{eq:output}, this
would reveal $S_0$ and $S_1$, from which the increment $c$ could be easily
deduced. This would take $2^{128}$ iterations of a very simple procedure, but is
completely infeasible.

The global strategy to obtain a practical prediction algorithm is essentially
the same as before (target a truncated version of the underlying linear
congruential generator), but is a bit more complicated to implement.

Set $\Delta S_i = S_{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
unknown increment $c$). Let us set:
\begin{align*}
  \nabla S_i \stackrel{def}{=} S_i - S_0 \equiv \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{align*}
Note that $\nabla S_0 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
\item Fully reconstruct $\Delta S_0$ from this partial knowledge and the outputs.
\item Reconstruct $S_0$ from $\Delta S_0$ and the outputs.
\end{enumerate}

@@

Another problem is that of confirming that the guesses are correct. When $c$ was
known, we could reconstruct the internal state; from there, filtering out the
bad guesses was easy. When $c$ is unknown, the same strategy does not work, but
a very strong consistency check can still be implemented. The essential idea is
the following: once $\nabla S_0[0:64+\ell]$ is known, then all the rotations can
be determined accurately and with little computational effort. This in turn
allows a consistency check between the partial information deduced from the
guesses and the actual outputs.

Once we have found $\nabla S_0[0:64+\ell]$, we can compute
$\nabla S_i[0:64+\ell]$ for any $i$ (geometric progression); because we have
guessed the first rotation and the $\ell$ least significant bits of the state,
using~\eqref{eq:output} we gain access to $S_0[58:64+\ell]$; combined with the
``differences'' $\nabla S_i$, this reveals $S_i[58:64+\ell]$ for any $i$ (and we
already had $S_i[0:\ell]$). This allows us to compute
$S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.

@@

More precisely, let us assume that we have guessed the $\ell$ least-significant
bits of $S_0$ (we denote them by $w_0$) and the first rotation $r_0$. Set
$Y_0 \stackrel{def}{=} X_0 \lll r_0$; there is an unknown value
$x_0$ such that:
\[
  S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
\]
The unknown value $x_0$ will not be recovered until the third phase of the whole
procedure. We obtain the $i$-th state by
$S_i \equiv \nabla S_i + S_0 \bmod 2^{64+\ell}$; however, because the ``middle''
of $S_0$ is unknown, an unknown \emph{carry} may cross the 64-th during the
addition and perturb $S_i[64:64+\ell]$. As a result, there are unknown
quantities $0 \leq y_i < 2^{64-\ell}$ and $0 \leq z_i \leq 1$ such that:
\begin{align*}
  S_i[0:\ell] &\equiv \nabla S_i + w_0 \bmod 2^\ell \\
  S_i[\ell:64] &\equiv y_i \\
  S_i[64:64+\ell] &\equiv z_i + \nabla S_i[64:64+\ell] + (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
\end{align*}

@@

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
  \State $T'_j \gets (T_j -  S'_i[58:64+\ell]) \bmod 2^{6+\ell}$ \Comment{Shifted truncated LCG}
  \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot \Delta U_0[0:58-\ell]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}
\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
\label{algo:unknown_1}
\end{algorithm}

@@

% \subsection{preuve cas difficile}
% \todo{À ranger au bon endroit}
% % Babai ça marche pas (genre zero)

% The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero.

% We call \(U\) the truncated vector : \(S'[\ell,128]\). We notice that \(U\) is in \(L(A,2^{128-\ell},n)\).

% We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the 6 upper bits of \(U\).

% (maybe dessin)

% We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use an exact CVP solver as explain in  (insererlasection) on \(T'\) and \(L(A,2^{128-\ell},n)\).

% As \(U\) and \(T'\) have exactly the same \(6\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(128-6-\ell)}}\).

% So, to be sure that our algorithm return the vector \(U\), we need to find sets of parameters \((n,\ell)\) such that :

% \[2\sqrt{n}^{(128-6-\ell)} \leqslant \lVert SVP(L(A,2^{128-\ell},n))\rVert \]

% \begin{itemize}
% 	\item For \(\ell=11\), the smallest \(n\) we need is 39.

% 	\item For \(\ell=12\), the smallest \(n\) we need is 38.

% 	\item For \(\ell=13\), the smallest \(n\) we need is again 38 .

% \end{itemize}

>>>>

    \draw (0.3, 1.1) -- (0.3, 1.5) -- (-1, 1.5) -- (-1, -4) -- node[above] {$r_i$} (5.8, -4);
    \draw (-1.1, -2.9) -- +(0.2, -0.2);
    \path (-0.9, -3) node[anchor=west] {6};

    \draw[->] (x) -- node[right] {$Y_i$} +(0, -1.5);
    \node at (6.4, -6.25) (xi) {$X_i$};
    \draw[->] (6.4, -4.4) -- (xi);

@@

This is a recurring theme is this paper, so we now state the problem in
general. If $a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there
is a \emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
$0 \leq 1 \leq \beta$) such that:
\begin{align}
  (a \boxplus b)[i:j]  &= a[i:j] \boxplus  b[i:j] \boxplus  \gamma, \label{eq:caryy}  \\
  (a \boxminus b)[i:j] &= a[i:j] \boxminus b[i:j] \boxminus \beta.  \label{eq:borrow}
\end{align}

@@

Let $\LCG_{k}(x, c)$ denote the vector $(u_0, u_1, u_2, \dots)$ of integers
modulo $2^k$.  It is easy to check that:
\begin{align}
\label{eq:lcg-additive}
\LCG_{k}(x + y, c + d) &= \LCG_{k}(x, c) + \LCG_{k}(y, d),  \\
\label{eq:lcg-scalar}
\LCG_{k}(\lambda x, \lambda c) &= \lambda \LCG_{k}(x, c).   
\end{align}
Hence those sets of vector have a lot of structure we can use.

The LCG have been largely studied. \todo{expand this} This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}. A particular LCG is the Lehmer's generator \cite{Lehmer}, where the multiplier \(a\) is known and the increment is zero.

@@

\begin{itemize}
	\item By \eqref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
	\item By \eqref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
\end{itemize}

@@

\[
  \begin{pmatrix}
    1 & a   & a^2 & \dots & a^{n-1} \\
    0 & 2^k & 0   & \dots & 0 \\
    0 & 0   & 2^k & \dots & 0 \\
    \dots & \dots & \dots & \dots & \dots\\
    0 & 0 & 0 & \dots & 2^k \\
  \end{pmatrix}
\]

@@

\todo{NOTE: J'introduis $\ell$ ici. ATTENTION RETENUES MODULAIRES.}  We want to
find an element \(U\) in the lattice only knowing the $\ell$ most significant
bits of each of its coordinates.  More precisely, let $T_i = U_i[k-\ell:k]$
denote the top $\ell$ bits of $U_i$. We assume that there is a ``noise vector''
$N$ such that $N_i \in \{-1, 0, 1\}$, and we set
$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$. The problem we have is the
following: recover $U$ from its noisy truncation $\widetilde{T}$.

In principle, $2^{k-\ell} \widetilde{T}$ should be ``close'' to a Lattice
point. \todo{Mais il faut se dire que $2^{k-\ell} T$ et
  $2^{k-\ell} \widetilde{T}$ peuvent être éloignés, surtout si $T_i = 0$ et
  $N_i=-1$, à cause du $\bmod 2^\ell$... L'argument de Julia était que dans ce
  cas-là, $2^{k-\ell} \widetilde{T}$ est proche d'un \emph{autre} point du
  réseau, en ajoutant un $2^k$ sur la $i$-ème composante.}

Decompose $U = T + (T-U)$ where $T$ contains only the upper bits and $T-U$ the lowers one. There is no reason for \(T\) to be in the lattice, but as \(T-U\) is small, we know \(T\) is \emph{close} to \(U\).

@@

\todo{En fait, on va juste utiliser $k=128$, donc établir $n=42$ est suffisant. Faire disparaître $\ell$.}
In this article, we will only use an exact CVP solver when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator. Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{122-\ell} \). So, if we fix \(\ell\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122-\ell} \leqslant \lambda_1(L(128-\ell,n))\].

@@

As using an exact CVP solver is costly, we will try a cheaper but less precise way to find the closest vector.

We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
\[\rVert G \lVert =  \sup_{x \in \mathbb{R}^n}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
In the case of the \(\lVert \cdot \rVert_2\) norm,
\(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular
value of $G$ (equivalently, $\sigma(G)$ is the absolute value of the largest
eigenvalue of $G^{t} G$ --- \todo{recheck this}).

@@

\todo{Ce serait beaucoup plus naturel à mon avis de dire qu'on connaît les $\ell$ bits de poids fort. A ce stade, rien ne vient justifier le ``+6''.}
In this article, we will only use the Babai's rounding when \(k = 64\) and when we know the \(6+\ell\) upper bit of our Lehmer's generator. Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{58-\ell} \). So, if we fix \(n\) we can search for the minimum \(\ell\) such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{58-\ell} \leqslant \lambda_1(L(n,64))\]

\begin{table}[htbp!]
  \centering
  \begin{tabular}{|c|c|c|c|}
	\hline
	n & \(\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert\) & \( \lambda_1(L(n,64)) \) & minimum \(\ell\) \\
	\hline
	3 & 2.87 & \(4.09e^{12} \simeq 2^{41.9}\) & 19 \\
	4 & 2.06 & \(2.44e^{14} \simeq 2^{47.8}\) & 13 \\
	5 & 3.77 & \(1.72e^{15}\simeq 2^{50.6}\) & 11 \\
	6 & 2.69 & \(1.03e^{16}\simeq 2^{53.2}\) & 8 \\
	\hline
  \end{tabular}
  \caption{\todo{WRITE-ME}}
  \label{tab:babai}
\end{table}

@@

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations $r_i$ give access to $S_i[58:64+\ell]$ for the corresponding
states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
linear congruential generator on $64$ bits, where we have access to the most
$6+\ell$ bits of each state (denoted by $T$), for a few consecutive states. This
is sufficient to reconstruct entirely the successive states of this truncated
linear congruential generator. This reveals $S_0[\ell:64+\ell]$, and
using~\eqref{eq:output} the entire $S_0$ can be reconstructed. The precise
details follow.

@@

We consider the sequence of internal states
$\mathbf{S} = (S_0, S_1, \dots) = \LCG_{128}(S_0, c)$. We will guess the $\ell$
least-significant bits of $S_0$, therefore let us assume that their value is
known and denote it by $w$. We define $\mathbf{S}' = \LCG_{128}(S_0 - w, 0)$ and
$\mathbf{K} = \LCG_{128}(w, c)$ (this is Known). By~\eqref{eq:lcg-additive}, we
have $\mathbf{S}' = \mathbf{S} - \mathbf{K}$. The point is that the elements of
$\mathbf{S}'$ follow a geometric progression of common ratio $a$; in addition,
the $\ell$ least significant bits of each components are equal to zero. It
follows that $\mathbf{S}'[\ell:64+\ell]$ follows a geometric progression of
common ratio $a$, modulo $2^{64}$. The crux of the reconstruction algorithm is
to find $\mathbf{S}'[\ell:64+\ell]$.

We know $K_i[58:64+\ell]$ for all $i$, and for each guessed rotation $r_i$ we
have access to $T_i \stackrel{def}{=} S_i[58:64+\ell]$.  We want
$T'_i \stackrel{def}{=} (S_i \boxminus K_i)[58:64+\ell]$, which is the
truncation of $S'_i$. Thanks to~\eqref{eq:borrow}, we know that there is an
unknown vector $\mathbf{B}$ of \emph{borrows}, whose components are either 0 or
1, such that
$\mathbf{T'} = \mathbf{S}[58:64+\ell] \boxminus \mathbf{K}[58:64+\ell] \boxminus
\mathbf{B}$.

Because the borrow are unknown, we in fact compute
$\widetilde{\mathbf{T}'} = \mathbf{S}[58:64+\ell] \boxminus
\mathbf{K}[58:64+\ell]$, and clearly
$\widetilde{\mathbf{T}'} = \mathbf{T}' \boxplus \mathbf{B}$.

We will guess $n$ rotations and $\ell$ least-significant bits of the state, for
a total of $2^{6n + \ell}$ guessed bits. Table~\ref{tab:babai} gives a
lower-bound on $\ell$ given $n$, and we see that the total number of guessed
bits reaches a minimum of $37$ when $n=3$ and $\ell=19$. Therefore, we guess
$\ell = 19$ low-order bits of the state and three consecutive rotations.

The algorithm that reconstructs the internal state of the \textsf{PCG64}
generator with known increment proceeds as shown in
algorithm~\ref{algo:known}. The point is that when the guesses are correct, then
from the truncated geometric series $\mathbf{T'}$, the solution of the CVP
instance reveals $U_j = S_j'[\ell:64+\ell]$. From there, the correction of the
algorithm is easily established.

The procedure is completely practical. More details are given in section~\ref{sec:implem}.

\begin{algorithm}
\begin{algorithmic}[1]
  \Procedure{ReconstructState${}_\ell$}{$X_0, X_1, X_2$}
  \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2$.}
  \State $\ell \gets 19$
  \For{$0 \leq w < 2^{\ell}$} \Comment{Guess least-significant bits of $S_0$}
%    \State $w_j = a^j w \bmod 2^{\ell}$ \Comment{Propagate to $S_1, S_2$}
    \State $K_j \gets a^j w + c (a^j - 1)(a-1)^{-1} \bmod 2^{128}$    \Comment{Known part}

@@

    \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$  \Comment{Truncated LCG output}

    \State $\widetilde T'_j \gets T_j \boxminus K_j[58:64+\ell]$  \Comment{Truncated geometric series}
    \State $(U_0, U_1, U_2) \gets \left\lfloor 2^{58-\ell} \cdot (\widetilde T'_0, \widetilde T'_1, \widetilde T'_2) \cdot  G_3^{-1} \right\rceil \cdot G_3$ \Comment{CVP (Babai rounding)}

    \State $S_0[0:64] \gets  K_0[0:64] + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
    \State $S_0[64:128] \gets S_0[0:64] \oplus Y_0$

    \State $S_1 \gets a  S_0 + c$ \Comment{Recompute $X_1$}
    \State $\widehat{Y}_1 =  S_1[0:64] \oplus  S_1[64:128]$

    \If{$\widehat {Y}_1 = Y_1$} \Comment{Check consistency}
    \State output $S_0$ as a candidate internal state.
    \EndIf
    \EndFor
    \EndFor
    \EndProcedure
  \end{algorithmic}
  \caption{State reconstruction Algorithm (case where $c$ is known)}
  \label{algo:known}
\end{algorithm}

@@

The algorithm of section~\ref{sec:Cknown} does not apply directly to the general
case where the value of $c$ is unknown. A ``baseline'' procedure would consist
in guessing $S_0[64:128]$ and $S_1[64:128]$; using eq.~\eqref{eq:output}, this
would reveal $S_0$ and $S_1$; from there, the increment $c$ is easily to obtain,
and every secret information has been reconstructed. This would take $2^{128}$
iterations of a very simple procedure, which is completely infeasible.

% The global strategy to obtain a practical prediction algorithm is essentially
% the same as before (target a truncated version of the underlying linear
% congruential generator), but is a bit more complicated to implement.

Set $\Delta S_i = S_{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
unknown increment $c$). The global ``guess-and-determine'' strategy is
essentially the same as before: gaining access to a truncated version of
$\Delta S_i$, solving a small SVP instance, reconstructing $\Delta S_0$, then
checking consistency.

Let us set:
\begin{align*}
  \nabla S_i \stackrel{def}{=} S_i \boxminus S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{align*}
Note that $\nabla S_1 = \Delta S_0$. The prediction algorithm we propose proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
\item Fully reconstruct $\Delta S_0$ from this partial knowledge and the outputs.
\item Reconstruct $S_0$ from $\Delta S_0$ and the outputs.
\end{enumerate}

@@

Confirming that the guesses are correct is less immediate. When $c$ was known,
we could reconstruct the internal state; from there, filtering out the bad
guesses was easy. When $c$ is unknown, the same strategy does not work, but a
very strong consistency check can still be implemented. The essential idea is
the following: once $\Delta S_0[0:64+\ell]$ is known, then all the rotations can
be determined accurately and with little computational effort. This in turn
allows a consistency check between the partial information deduced from the
guesses and the actual outputs.

Once we have found $\Delta S_0[0:64+\ell]$, we can compute
$\nabla S_i[0:64+\ell]$ for any $i$ (geometric progression); because we have
guessed the first rotation and the $\ell$ least significant bits of the state,
using~\eqref{eq:output} we gain access to $S_0[58:64+\ell]$; combined with the
``differences'' $\nabla S_i$, this reveals $S_i[58:64+\ell]$ for any $i$ (and we
already had $S_i[0:\ell]$). This allows us to compute
$S_i[0:\ell] \oplus S_i[64:64+\ell]$ for any $i$.

@@

More precisely, let us assume that we have guessed the $\ell$ least-significant
bits of $S_0$ (we denote them by $w_0$) and the first rotation $r_0$. Set
$Y_0 \stackrel{def}{=} X_0 \lll r_0$; there is an unknown value
$x_0$ such that:
\[
  S_0 = w_0 + 2^\ell x + 2^{64} (w_0 \oplus Y_0[0:\ell]) + 2^{64+\ell} (x \oplus Y_0[\ell:64]) \qquad \left(0 \leq x < 2^{64-\ell}\right).
\]
The unknown value $x_0$ will not be recovered until the third phase of the whole
procedure. We obtain the $i$-th state by $S_i \equiv \nabla S_i \boxplus S_0$;
however, because the ``middle'' of $S_0$ is unknown, we have the same problem as
in eq.~\eqref{eq:diff_borrow}: an unknown \emph{carry} may cross the 64-th bit
during the addition and perturb $S_i[64:64+\ell]$. As a result, there is an
unknown vector $\mathbf{C}$, whose components are either 0 or 1, such that such
that:
\begin{align*}
  S_i[64:64+\ell] &= C_i \boxplus \nabla S_i[64:64+\ell] \boxplus (w_0 \oplus Y_0[0:\ell]) \bmod 2^\ell \\
\end{align*}

@@

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
  \State $K_j \gets a^j w_0 + (2c+1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
  \State $T'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Truncated geometric series}
  \State \todo{Attention double retenue : $\pm 1$.}
  \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot \Delta U_0[0:58-\ell]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}
\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
\label{algo:unknown_1}
\end{algorithm}

@@
